---
title: "Iris Dataset Statistics Lab"
author: "Alex Hurtado Desaulniers"
output:
  html_notebook: default
  pdf_document: default
---

# Libraries
```{r}
library(tidyverse)
```

# Load the iris dataset


```{r}
data(iris)
```

```{r} 
# Changing names of each column
colnames(iris) <- c("SepalLen", "SepalWid", "PetalLen", "PetalWid", "Species")
```

```{r} 
# checking  first few rows
head(iris)
```

## Questions:
 1. How many numeric columns are in the dataset? 4
 2. How many species are included? 3
 3. What is the range of SepalLen? (min and max) Minimum is 4.6 Maximum is 5.4 giving a range of 0.8.

# Summary Statistics
Calculate mean, median, variance, SD for numeric columns

```{r}
sapply(iris[,1:4], mean)
sapply(iris[,1:4], median)
sapply(iris[,1:4], var)
sapply(iris[,1:4], sd)
```

## Questions:
1. Which variable has the largest spread? 
PetalLen has the largest spread, with a variance of 3.116 and a standard deviation of 1.765.
2. Which has the smallest spread?
SepalWid has the smallest spread, with a variance of 0.190 and a standard deviation of 0.436.
3. How does the mean compare to the median for PetalLen?
The mean for PetalLen is 3.758, which is lower than the median of 4.35
4. What does a large variance indicate?
A large variance indicates that the data points are spread out further from the mean, suggesting high variability in those measurements.

# Summary by Species

```{r}
aggregate(SepalLen ~ Species, data=iris, FUN=mean)
aggregate(PetalWid ~ Species, data=iris, FUN=var)
```

## Questions:
 1. Which species has the largest SepalLen on average? virginica 
 2. Which species shows the most variability in PetalWid? virginica, with a variance of 0.0754.
 3. What does this tell you about the characteristics of different species?
It tell me that some that species are not uniform and that some like virginica may grow larger or have more physical diversity in certain traits than others.

# Visualizations

### Histogram of SepalLen

```{r}
hist(iris$SepalLen,
     main="Sepal Length Distribution",
     xlab="Sepal Length (cm)",
     col="pink",
     border="black")
```

## Questions:
 1. What is the shape of the distribution (symmetric, skewed)?
 The distribution appears slightly left-skewed as there is a shorter "tail" toward the higher values (7–8 cm)
 2. Are there any unusually low or high values?
Yes 4,7, and 8 are unusually low comapared to the rest of the graph.

# Boxplot of PetalLen by Species

```{r}
boxplot(PetalLen ~ Species,
        data=iris,
        main="Petal Length by Species",
        ylab="Petal Length (cm)",
        col=c("pink","skyblue","purple"))
```

## Questions:
 1. Which species has the longest petals? virginica
 2. Are there any outliers visible? Yes, there are visible outliers (the small circles) for both setosa and versicolor.
 3. How do the distributions differ between species?
Setosa has very short, tightly clustered petals; versicolor is in the middle range; and virginica has the longest petals with the most spread.

# Correlation Matrix

```{r}
cor_matrix <- cor(iris[,1:4])
cor_matrix
```

## Questions:
 1. Which two variables are most strongly correlated? 
 PetalLen and PetalWid have the strongest positive correlation at 0.963.
 2. Are there any weak or negative correlations?
 Yes, SepalLen and SepalWid have a weak negative correlation (-0.118), and SepalWid is negatively correlated with both petal measurements.
 
 3. How might these correlations help distinguish species?
Strong correlations between petal length and width suggest that as one grows, the other does too, creating a specific "shape" profile that can be used to identify a species.

# Interpretation 

Write a short paragraph (3-5 sentences) answering:

- Which variables best distinguish species?
- How could a scientist use these statistics in real life?
- Do the histograms, boxplots, and correlations tell a consistent story?

PetalLen and PetalWid are the best variables for distinguishing species because their distributions (as seen in the boxplots) show almost no overlap between setosa and the other groups. A scientist could use these statistics in real life to develop automated classification tools for botany or to study evolutionary trends in plant morphology. The visualizations and correlations tell a consistent story: petal dimensions are highly predictable and serve as the primary distinguishing physical characteristic between these three Iris species.
